---
title: "SQL"
format: html
author: "Ryan Mooney"
---
# The task
For today's project, I will be working with data from the [Stanford Open Policing Project](https://openpolicing.stanford.edu/) pubished by Pierson, et al. 2020. “[A Large-Scale Analysis of Racial Disparities in Police Stops Across the United States](https://www-nature-com.ccl.idm.oclc.org/articles/s41562-020-0858-1).” Nature Human Behaviour, 1–10. I will be investigating a few of the many tables contained within the database.

First, let's set up the connection to the traffic database.

```{r}
con_traffic <- DBI::dbConnect(
  RMariaDB::MariaDB(),
  dbname = "traffic",
  host = Sys.getenv("TRAFFIC_HOST"),
  user = Sys.getenv("TRAFFIC_USER"),
  password = Sys.getenv("TRAFFIC_PWD")
)
```

```{r}
#| warning: FALSE
#| message: FALSE
library(tidyverse)
library(DBI)
```


Let's have a look at all of the tables that are in the traffic database, so we know what we are working with.
```{r}
#| connection: con_air
DBI::dbListTables(con_traffic)
```

```{sql}
#| connection: con_traffic

SELECT * FROM co_statewide_2020_04_01 LIMIT 8;
```

As I am from Colorado, I am curious to look more specifically at the state and how vehicular stops compare between some states I might want to live in (namely, Colorado, California, and New York). 

First, I'd like to make a plot to show how vehicular stops have changes over the years in these three states.

```{sql}
#| connection: con_traffic

SELECT
    YEAR(date) AS year,
    COUNT(*) AS total_stops
FROM co_statewide_2020_04_01
WHERE type = 'vehicular'
GROUP BY YEAR(date)
ORDER BY YEAR(date);
```
```{r}
co_stops_per_year <- dbGetQuery(
  con_traffic,
  "
  SELECT
      YEAR(date) AS year,
      COUNT(*) AS total_stops
  FROM co_statewide_2020_04_01
  WHERE type = 'vehicular'
  GROUP BY YEAR(date)
  ORDER BY YEAR(date);
  "
)
```

```{r}
#| warning: FALSE
#| message: FALSE
ggplot(co_stops_per_year, aes(x = year, y = total_stops)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Total Vehicular Stops in Colorado per Year",
    x = "Year",
    y = "Number of Stops"
  )
```
Now, I am going to try to UNION all three tables (co_statewide, ca_statewide, and ny_statewide), so that I can see how the states differ in their number of vehicular stops over the past few years. 

```{r}
co_ca_ny_stops_per_year <- dbGetQuery(
  con_traffic,
  "
  SELECT
      'Colorado' AS state,
      YEAR(date) AS year,
      COUNT(*) AS total_stops
  FROM co_statewide_2020_04_01
  WHERE type = 'vehicular'
  GROUP BY state, YEAR(date)

  UNION ALL

  SELECT
      'California' AS state,
      YEAR(date) AS year,
      COUNT(*) AS total_stops
  FROM ca_statewide_2023_01_26
  WHERE type = 'vehicular'
  GROUP BY state, YEAR(date)

  UNION ALL

  SELECT
      'New York' AS state,
      YEAR(date) AS year,
      COUNT(*) AS total_stops
  FROM ny_statewide_2020_04_01
  WHERE type = 'vehicular'
  GROUP BY state, YEAR(date)

  ORDER BY year, state;
  "
)
```

```{r}
#| warning: FALSE
#| message: FALSE

ggplot(co_ca_ny_stops_per_year, aes(x = year, y = total_stops, color=state)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Total Vehicular Stops in Colorado per Year",
    x = "Year",
    y = "Number of Stops"
  )
```


# References
Pierson, E., Simoiu, C., Overgoor, J. et al. A large-scale analysis of racial disparities in police stops across the United States. Nat Hum Behav 4, 736–745 (2020).