---
title: "UNESCO"
format: html
---

```{r}
#| message: false
#| warning: false
library(tidyverse)
```

# The Data

I wanted to explore a simple data set comparing UNESCO world heritage sites in Norway, Denmark, and Sweden. The original data is from [UNESCO](https://whc.unesco.org/en/list) and was downloaded from [TidyTuesday](https://github.com/rfordatascience/tidytuesday/blob/main/data/2024/2024-02-06/readme.md). The data also comes from the {pixarfilms} R package by [Eric Leung](https://github.com/erictleung).


```{r}
#| message: false
#| warning: false
unesco <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2024/2024-02-06/heritage.csv')
```

```{r}
unesco2 <- unesco |>
  pivot_longer(
    cols = c(`2004`, `2022`),
    names_to = "year",
    values_to = "number_of_sites"
  )
```

# Let's plot the data!

```{r}
ggplot(unesco2, aes(x = country, y = number_of_sites, color = year)) +
  geom_point(size = 3) +
  labs(
    title = "UNESCO World Heritage Sites",
    x = "Country",
    y = "Number of Sites"
  )
```

Showing the number of World Heritage Sites in Norway, Denmark, and Sweden across 2004 to 2022.

Image credit: Ryan Mooney
